---
title: "study_guide"
output: html_document
date: "2024-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### what is in each notes 

## lab 4 is Dplyr
### using slecet to pull out specific variable 
```{r}
select(fish, "lakeid", "scalelength")
```

To add a range of columns use `start_col:end_col`.
```{r}
select(fish, fish_id:length)
```

The - operator is useful in select. It allows us to select everything except the specified variables.
```{r}
select(fish, -"fish_id", -"annnumber", -"length", -"radii_length_mm")

For very large data frames with lots of variables, `select()` utilizes lots of different operators to make things easier. Let's say we are only interested in the variables that deal with length.

```{r}
names(fish)
```

```{r}
select(fish, contains("length"))
```

When columns are sequentially named, `starts_with()` makes selecting columns easier.
```{r}
select(fish, starts_with("radii"))
```

Options to select columns based on a specific criteria include:  
1. ends_with() = Select columns that end with a character string  
2. contains() = Select columns that contain a character string  
3. matches() = Select columns that match a regular expression  
4. one_of() = Select columns names that are from a group of names  

```{r}
names(fish)
```

```{r}
select(fish, ends_with("id"))
```

```{r}
select(fish, contains("fish"))

To select all columns that are *not* a class of data, you need to add a `~`.
```{r}
select_if(fish, ~!is.numeric(.)) #the ! means not
### Ways to use select


### ways to use filter 
## `filter()`
Unlike `select()`, `filter()` allows us to extract data that meet specific criteria within a variable. Let's say that we are interested only in the fish that occur in lake "AL". We can use `filter()` to extract these observations.  
```{r}
filter(fish, lakeid == "AL")
```

Similarly, if we are only interested in fish with a length greater than or equal to 350 we can use `filter()` to extract these observations.  
```{r}
filter(fish, length >= 350)
```

+ `filter()` allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal).  

Using the `!` operator allows for the exclusion of specific observations.
```{r}
not_AL <- filter(fish, lakeid != "AL") #! means not
```

## Using `filter()` with multiple observations  
Filtering multiple values within the same variable requires the `%in%` [operator](https://www.tutorialspoint.com/r/r_operators.htm).    
```{r}
filter(fish, length %in% c(167, 175))
```

Alternatively, you can use `between` if you are looking for a range of specific values.  
```{r}
filter(fish, between(scalelength, 2.5, 2.55))
```

You can also extract observations "near" a certain value but you need to specify a tolerance.  
```{r}
filter(fish, near(radii_length_mm, 2, tol = 0.2))
```

## Using `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. Below we extract only the fish that have lakeid "AL" and length >350.
```{r}
filter(fish, lakeid == "AL" & length > 350)
```

Notice that the `|` operator generates a different result.
```{r}
filter(fish, lakeid == "AL" | length > 350)
```

Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met.  
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 or condition 2 is met.  
+ `filter(xor(condition1, condition2)` will return all rows where only one of the conditions is met, and not when both conditions are met.  

In this case, we filter out the fish with a length over 400 and a scale length over 11 or a radii length over 8.
```{r}
filter(fish, length > 400, (scalelength > 11 | radii_length_mm > 8))
```

In this case, we filter out the fish with a length over 400 and a scale length over 11 or a radii length over 8.
```{r}
filter(fish, length > 400, (scalelength > 11 | radii_length_mm > 8))
### How to change the the class of a
variable 


```{r}
elephants$sex <- as.factor(elephants$sex)
```

### How to clean names
```{r}
elephants <- clean_names(elephants)
```
### How to load data
```{r}
read.csv("data/whatever.csv")
```

### how to pick out specific numbers from data 
#ex:how many males and females 
```{r}
elephants %>% 
  count(sex)
```

### How to take averages 
#ex:What is the average age all elephants in the data?
```{r}
elephants %>% 
  summarize(mean_age=mean(age))
```
## Adding Rows
We can add a new bird's data with the `rbind()` function, known as row bind. It will bind our new vector to our data frame row-wise.
```{r}
new_bird <- c("female", 3.6, 3.9)
new_bird
```

```{r}
hbirds<- rbind(hbirds, new_bird)
hbirds
```

## Adding Columns
What if we come across more data and we want to add it to our data frame? We can also do that with the `$` operator.
```{r}
hbirds$neighborhood <- c("lakewood", "brentwood", "lakewood", "scenic Heights")
hbirds
```

## Summary functions
Once data have been uploaded, you may want to get an idea of its structure, contents, and dimensions. I routinely run one or more of these commands when data are first imported.  

We can summarize our data frame with the`summary()` function.  
```{r}
summary(fish)
```

`glimpse()` is another useful summary function.
```{r}
glimpse(fish)
```

`nrow()` gives the numbers of rows.
```{r}
nrow(fish)
```

`ncol` gives the number of columns.
```{r}
ncol(fish)
```

`dim()` gives the dimensions.
```{r}
dim(fish)
```

`names` gives the column names.
```{r}
names(fish)
```

`head()` prints the first n rows of the data frame.
```{r}
head(fish)
```

`tail()` prinst the last n rows of the data frame.
```{r}
tail(fish)
```

`table()` is useful when you have a limited number of categorical variables. It produces fast counts of the number of observations in a variable. We will come back to this later... 
```{r}
table(fish$lakeid)
```

We can also click on the `fish` data frame in the Environment tab or type View(fish).
```{r}
View(fish)
```

## Filter
Filter is a way of pulling out observations that meet specific criteria in a variable. We will work a lot more with this in the next lab.  
```{r}
little_fish <- filter(fish, length<=100)
little_fish
```

## Practice
1. How many genera are represented in the msleep data frame?
```{r}
msleep %>% 
  summarize(n_genera=n_distinct(genus))
```

Here is another option.
```{r}
msleep %>% 
  tabyl(genus)
```