---
title: "BIS15L_AI_hmwk"
author: "Adara LoCascio"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
This homework is a bit less traditional this week. There are some errors in the code below. In addition add notes and titles to the code chunks that explains what the code chunck does. Use AI tools, notes from class, or any other resources to help fix the errors. Feel free to edit any plots to your preference. Don't forget to answer the two questions at the bottom. Upon completion, you should not have to '#' out any of the code chunks.

Push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.   

Again make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  


```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```


1.
```{r}
blood_count <- read.csv("data/blood_count_dataset.csv") %>% clean_names()
```

2. 
```{r}
glimpse(blood_count)

```

3. 
```{r}
blood_count %>% 
  group_by(gender) %>% 
  summarize(min=min(red_blood_cells),
            mean=mean(red_blood_cells),
            max=max(red_blood_cells))
```
```{r}
options(scipen = 999, digits = 2)
```


4. 
```{r}
blood_count %>% 
  filter(gender == "male") %>%
  select(3:9) %>%
  colMeans(na.rm = TRUE)

```

5. 
```{r}
blood_count %>% 
  filter(gender == "Female") %>% 
  select(3:9) %>%
  colMeans(na.rm = T)
```


6. 
```{r}

blood_count %>%
  pivot_longer(cols = 3:9, names_to = "test", values_to = "values") %>% 
  ggplot(aes(x = gender, y = values, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~ test, scales = "free") + 
  labs(title = "Boxplot of Blood Count Tests",
       x = "Test",
       y = "Values")

```

7. 
```{r}
grouped_bc <- blood_count %>%
  mutate(Age_Group = case_when(
    age >= 20 & age <= 29 ~ "20s",
    age >= 30 & age <= 39 ~ "30s",
    age >= 40 & age <= 49 ~ "40s",
    age >= 50 & age <= 59 ~ "50s",
    age >= 60 & age <= 70 ~ "60s"))

```

8. 
```{r}
grouped_bc %>% 
  pivot_longer(cols = 3:9, names_to = "test", values_to = "values") %>%
  ggplot(aes(x = Age_Group, y = values, fill = Age_Group)) + 
  geom_boxplot() +
  facet_wrap( ~test, scales = "free") +  
  labs(title = "Boxplot of Blood Count Tests by Age Group",
       x = "Age Group",
       y = "Values")

```
9. What are two drawbacks and two benefits of AI?


Benefits: AI streamlines tasks and enables data-driven decision-making, enhancing efficiency and outcomes across various fields.

Drawbacks: Job displacement and biases in AI algorithms pose challenges, potentially exacerbating unemployment and perpetuating inequalities if not addressed.

10. Do you think you will use AI for coding in the future? Why or why not?

Yes, I believe AI will be an integral part of coding in the future. AI can automate repetitive tasks, optimize code, and provide insights that can improve the quality and efficiency of coding.
## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences. 